The Astropy Project: \\ Sustaining and Growing a Community-oriented Open-source Project and the Latest Major Release (v5.0) of the Core Package

### 2022-03-15 src/ms.tex
remove duplicate numpy ref

Overview and statistics of contributors. Highlight changes since v2.0.

\begin{figure}
    \begin{centering}
      % \includegraphics{figures/contributor-summary.pdf}
        \caption{Placeholder figure!}
        \label{fig:contributor-summary}
    \end{centering}
\end{figure}

@@

\begin{longrotatetable}
    \begin{deluxetable*}{ccp{2in}p{2.5in}}
    \tablecaption{Registry of affiliated packages.}
    \label{tab:registry}
    \tablehead{
        \colhead{Package Name} &
        % \colhead{Stable} &
        \colhead{PyPI Name} &
        \colhead{Maintainer(s)} &
        \colhead{Citation(s)}
      }
      \startdata
        \input{generated/affiliated-table.tex}
      \enddata
  \end{deluxetable*}
\end{longrotatetable}

>>>>

% TODO: move this subsection to "updates to core package" section above?

\begin{figure}[t!]
    \begin{centering}
      \includegraphics[width=0.9\textwidth]{Ncommitters_vs_frac_commits.pdf}
        \caption{
            The cumulative fraction of Git commits to the \astropypkg core
            package as a function of number of contributors (committers) broken
            into three equal-length time periods between the start of \astropy
            and the feature freeze date of v5.0.
            The time eras early, mid, and recent correspond to the approximate
            year ranges 2011--2015, 2015--2018, 2018--2021 (defined more
            precisely in Section~\ref{sec:project-contributors}).
            In all eras, the majority of commits to the \astropypkg core package
            have come from a small subset of the total number of contributors.
            For example, as annotated in the figure, 90\% of the Git commits
            have come from 18, 34, and 44 committers in the early, mid, and
            recent eras, respectively.
        }
        \label{fig:contributor-summary:Ncommitters}
    \end{centering}
\end{figure}

As described in the introduction above, the \astropy Project is broader than the
core \astropypkg package as it also represents a number of core infrastructure
packages, affiliated packages, educational initiatives, and user communities.
In total, well over 1500 people have contributed in some way to the Project,
whether, for example, through contributing code or documentation, opening issues
on \github, providing infrastructure support, writing tutorials, or helping to
coordinate workshops that relate to Astropy.
The vast majority of these contributors and participants have interacted as
volunteers choosing to engage with the open-source and open-development software
community.
However, while the total number of contributors is large, the number of
contributions and bulk of the maintenance of the \astropy Project components has
primarily been sustained by a small number of heavily-engaged or
explicitly-employed contributors \citep[see, e.g., ][]{astropy:2018}, as is the
case in many large, open-source software communities
\citep[e.g.,][]{numpy:nature}.

As an example of this within the \astropy Project, here we focus on
contributions made to the \astropypkg core package.
Figure~\ref{fig:contributor-summary:Ncommitters} shows the number of
contributors who have contributed a given fraction of the total number of Git
commits, separated into three separate time spans (``eras'') in the history of
the Project.\footnote{Note, however, that measuring these statistics in terms of
Git commits is not a perfect way of encapsulating ``effort,'' but we use this
metric for its simplicity and clear definition.}
The annotations show that, within the early (July 2011--January 2015), mid
(January 2015--June 2018), and recent (June 2018--November 2021) eras, 90\% of
the commits to the \astropypkg core package have come from 18, 34, and 44
contributors.
While the trend is encouraging --- a larger pool of contributors are
contributing more, fractionally --- this emphasizes that, still, the vast
majority of the work in \astropypkg is done my a small subset ($<20\%$) of the
115, 216, and 234 contributors, respectively, within each era.

\begin{figure}[t!]
    \begin{centering}
      \includegraphics[width=0.75\textwidth]{top_contributors_career.pdf}
        \caption{
            The career stages and/or positions of the top contributors within
            each time era (defined as having contributed $>50$ commits to the
            \astropypkg core package in each era).
            The time eras early, mid, and recent correspond to the approximate
            year ranges 2011--2015, 2015--2018, 2018--2021 (defined more
            precisely in Section~\ref{sec:project-contributors}).
        }
        \label{fig:contributor-summary:top}
    \end{centering}
\end{figure}

Since its inception, the most active contributors to the \astropypkg package
have spanned a range of career stages and career paths.
For example, Figure~\ref{fig:contributor-summary:top} shows a breakdown of the
career stage/path at the midpoint of the three time eras defined in the previous
paragraph for all contributors with more than 50 commits to the \astropypkg
package within each era.
Over the $\sim$6 years between the ``early'' and ``recent'' eras, many
contributors moved between these categories (i.e., have graduated as students to
become research staff / faculty or software engineers), but also a number of
early contributors moved on to other positions and stopped contributing to the
core \astropypkg package.
Given this, it is encouraging that the numbers in each career stage/path have
stayed relatively constant over the history of \astropy (with some showing
growth), as new, active contributors have generally joined and stepped in to
fill top contributor roles within the project.
On the other hand, the total number of top contributors has not grown
substantially, reflecting one of our principal challenges: Sustaining a pipeline
from users to contributors to maintainers of the package.
One other thing to note in this context is that in all eras, 50--60\% of the
software engineers who contributed to \astropypkg were (or are) employed by
STScI, whose leadership have consistently provided resources (financial and
personnel) towards sustaining \astropypkg.

\begin{figure}[t!]
    \begin{centering}
      \includegraphics[width=0.75\textwidth]{commits_committers_per_month.pdf}
        \caption{
            \textbf{Top panel:} The number of Git commits per month (defined as
            consecutive 30 day windows) merged into to the \astropypkg core
            package over the whole history of the project (up to end of 2021,
            v5.0).
            The dark line shows the mean rate using a 1 year rolling window, but
            the under-plotted gray line shows the raw monthly rate.
            The raw number clearly peaks just before feature freezes (vertical
            lines) as a flurry of activity leading up to each release, but in
            recent years the mean rate has remained steady at about $\sim$300
            commits per month.
            \textbf{Bottom panel:} Similar to above, but now showing the number
            of unique Git committers who contribute to the \astropypkg core
            package each month (again defined as consecutive 30 day windows).
            While this also shows some variation depending on proximity to
            feature freezes, in recent years the mean rate of unique
            contributors has remained consistent at about $\sim$22 committers
            per month.
        }
        \label{fig:contributor-summary:per-month}
    \end{centering}
\end{figure}

A final metric we consider as a diagnostic of the contributor base and
contributions to the \astropypkg core package is the rate of contributions and
contributor participation, defined again using Git commits to quantify
contributions.
Figure~\ref{fig:contributor-summary:per-month} shows the monthly rates of
commits and unique committers to the \astropypkg core package over the history
of the package.
After an initial ramp-up in the commit rate and number of committers (lasting
roughly from 2011 to 2014), the commit rate and number of unique committers have
both remained fairly constant.

@@

\begin{longrotatetable}
    \begin{deluxetable*}{ccp{2in}p{2.5in}}
    \tablecaption{Registry of affiliated packages.}
    \label{tab:affiliated-registry}
    \tablehead{
        \colhead{Package Name} &
        % \colhead{Stable} &
        \colhead{PyPI Name} &
        \colhead{Maintainer(s)} &
        \colhead{Citation(s)}
      }
      \startdata
        \input{generated/affiliated-table.tex}
      \enddata
  \end{deluxetable*}
\end{longrotatetable}
